\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-06-20 2-cubble.Rmd
edit on the ERA5 example

### Creating from separate spatial and temporal tables

@@

### Coercing from foreign objects

@@

```{r echo = TRUE, eval = FALSE}
climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
```

In \proglang{R}, there are several packages available for wrangling NetCDF data, including \code{ncdf4}, \code{RNetCDF}, and \code{tidync}. The code below converts a NetCDF object of class \code{ncdf4} into a cubble object:

```{r echo = TRUE}
path <- system.file("ncdf/era5-pressure.nc", package = "cubble")
raw <- ncdf4::nc_open(path)
as_cubble(raw)
```

Sometimes, analysts may choose to read onlny a subset of the NetCDF data. In such cases, the arguments \code{vars}, \code{long_range} and \code{lat_range} can be used to subset the data based on variable and the grid resolution:

```{r echo = TRUE, eval = FALSE}
as_cubble(raw, vars = "q",
          long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1))
```

>>>>

@@

@@

```{r echo = TRUE}
climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
```

### 2023-06-20 4-4-netcdf.Rmd
edit on the ERA5 example

The ERA5 data [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF) and can be downloaded from Copernicus Climate Data Store (CDS) [@cds] website or via the \pkg{ecmwfr} package [@ecwmfr]. This example reproduces the ERA5 data row of Figure 19 in @hersbach2020era5 to show the southern polar vortex splitting into two on 2002-09-26 and further splitting into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also in @simmons2020global and @simmons2005ecmwf.

For the reproduction, we focus on the \code{era5-pressure} data, hourly pressure levels from 1970 to present, with the *specific humidity* and *geopotential*. The downloaded NetCDF data (`raw`) is first converted to a \code{cubble} object:

<!--Here we specifically look at the information  contained on the variables *specific humidity* measuring the mass of water vapor per kilogram of moist air  and the *geopotential* variable recording the gravitational potential energy of a unit mass, at a particular location, relative to mean sea level.-->

<!--The goal of this example is to reproduce the break-up of the southern polar vortex that happened in late September and early October in 2002 in the stratosphere in the Southern Hemisphere. Therefore, we filter data corresponding to the Southern Hemisphere and restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04.--> 
```{r echo = FALSE, eval = TRUE}
raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))
```

```{r echo = TRUE}
dt <- as_cubble(
  raw, vars = c("q", "z"),
  long_range = seq(-180, 180, 1), lat_range = seq(-88, 88, 1))
```

<!-- Figure \ref{fig:netcdf} reproduces the ERA5 data row of Figure 19 in @hersbach2020era5. It shows the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04 in the stratosphere. Readers interested in the analysis of this figure can refer to @hersbach2020era5, @simmons2020global, and @simmons2005ecmwf for more details.-->

Creating the plot requires making transformations on time, unfolding the data for computing the statistic of interest, which is plotted directly as a contour plot with ggplot.

@@

```{r netcdf, out.width="100%", fig.height = 4, fig.width = 10, fig.cap = "A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002. The polar vortex, signalled by the high specific humidity, splits into two on 2002-09-26 and further splits into four on 2002-10-04."}
library(sf)
date <- c("2002-09-22", "2002-09-26", "2002-09-30", "2002-10-04") |> as.Date()
res <- dt |> 
  face_temporal() |> 
  filter(lubridate::date(time) %in% date) |>
  unfold(long, lat) |> 
  mutate(q =  q* 10^6)

res |> 
  ggplot() +
  # q for specific humidity
  geom_point(aes(x = long, y = lat, color = q), size = 2) + 
  #geom_tile(aes(x = long, y = lat, fill = q), color = "transparent") +
  # z for geopotential
  geom_contour(data = res, aes(x = long, y = lat, z = z),
               color = "grey20", binwidth = 4000, linewidth = 0.5) +
  geom_sf(data = country , alpha = 0.5, fill = "transparent", color = "lightgreen") +
  coord_sf(
    crs = "+proj=stere +lat_0=-90 +lon_0=-180 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs ", default_crs = sf::st_crs(4326), clip = "off") +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_blank()) +
  facet_wrap(vars(as.Date(time)), nrow = 1)+  
  colorspace::scale_color_continuous_sequential("Purple-Yellow", name = "Specific humidity") + 
  theme(legend.text = element_text(size = 13),
        legend.title = element_text(size = 13)) + 
  labs(x ="", y = "")
```

>>>>

The ERA5 reanalysis [@hersbach2020era5] provides hourly estimates of atmospheric, land and oceanic climate variables on a global scale. The hourly pressure level data in the NetCDF format can be downloaded from Copernicus Climate Data Store (CDS) [@cds] or via the \pkg{ecmwfr} package [@ecwmfr]. This example reproduces the row created from ERA5 reanalysis data shown in Figure 19 by @hersbach2020era5. The plot shows the southern polar vortex splitting into two on 2002-09-26, and further splitting into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also in @simmons2020global and @simmons2005ecmwf.

To work with NetCDF data in R, several packages are available, including \code{ncdf4}, \code{RNetCDF}, and \code{tidync}. The following code converts a NetCDF object of class \code{ncdf4} into a cubble object:

```{r echo = TRUE}
raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))
```

Analysts can extract a subset of the NetCDF data with the arguments \code{vars}, \code{long_range} and \code{lat_range}. In this example, the variables q (specific humidity) and z (geopotential) are selected and the coordinates are subsetted to every degree in longitude and latitude:

```{r echo = TRUE}
(dt <- as_cubble(raw, vars = c("q", "z"),
                 long_range = seq(-180, 180, 1), lat_range = seq(-88, -15, 1)))
```

Once the NetCDF data is coerced into a cubble object, subsequent analysis can be conducted to filter on the date of interest, scale the variable specific humidity and create visualisation in ggplot as in Figure \ref{fig:netcdf}.

```{r netcdf, out.width="100%", fig.height = 4, fig.width = 10, fig.cap = "A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002. The polar vortex, signalled by the high specific humidity, splits into two on 2002-09-26 and further splits into four on 2002-10-04."}
date <- c("2002-09-22", "2002-09-26", "2002-09-30", "2002-10-04") |> as.Date()
res <- dt |> 
  face_temporal() |> 
  filter(lubridate::date(time) %in% date) |>
  unfold(long, lat) |> 
  mutate(q = q* 10^6)

res |> 
  ggplot() +
  # q for specific humidity
  geom_point(aes(x = long, y = lat, color = q)) + 
  #geom_tile(aes(x = long, y = lat, fill = q), color = "transparent") +
  # z for geopotential
  geom_contour(data = res, aes(x = long, y = lat, z = z),
               color = "grey20", binwidth = 4000, linewidth = 0.5) +
  geom_sf(data = country , alpha = 0.5, fill = "transparent", color = "lightgreen") +
  coord_sf(
    crs = "+proj=stere +lat_0=-90 +lon_0=-180 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs", default_crs = sf::st_crs(4326), clip = "off") +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_blank()) +
  facet_wrap(vars(as.Date(time)), nrow = 1)+  
  colorspace::scale_color_continuous_sequential("Purple-Yellow", name = "Specific humidity") + 
  theme(legend.text = element_text(size = 13),
        legend.title = element_text(size = 13)) + 
  labs(x ="", y = "")
```

@@

### 2023-06-20 jss/main.tex
edit on the ERA5 example

\hypertarget{creating-from-separate-spatial-and-temporal-tables}{%
\subsubsection{Creating from separate spatial and temporal tables}\label{creating-from-separate-spatial-and-temporal-tables}}

@@

\hypertarget{coercing-from-foreign-objects}{%
\subsubsection{Coercing from foreign objects}\label{coercing-from-foreign-objects}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\end{CodeChunk}

In \proglang{R}, there are several packages available for wrangling NetCDF data, including \code{ncdf4}, \code{RNetCDF}, and \code{tidync}. The code below converts a NetCDF object of class \code{ncdf4} into a cubble object:

\begin{CodeChunk}
\begin{CodeInput}
R> path <- system.file("ncdf/era5-pressure.nc", package = "cubble")
R> raw <- ncdf4::nc_open(path)
R> as_cubble(raw)
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [26565], index: time, nested form
# spatial:  [113, -53, 153, -12], Missing CRS!
# temporal: time [dttm], q [dbl], z [dbl]
     id  long   lat ts              
  <int> <dbl> <dbl> <list>          
1     1  113    -12 <tibble [6 x 3]>
2     2  113.   -12 <tibble [6 x 3]>
3     3  114.   -12 <tibble [6 x 3]>
4     4  114.   -12 <tibble [6 x 3]>
5     5  114    -12 <tibble [6 x 3]>
# i 26,560 more rows
\end{CodeOutput}
\end{CodeChunk}

Sometimes, analysts may choose to read onlny a subset of the NetCDF data. In such cases, the arguments \code{vars}, \code{long_range} and \code{lat_range} can be used to subset the data based on variable and the grid resolution:

\begin{CodeChunk}
\begin{CodeInput}
R> as_cubble(raw, vars = "q",
+           long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1))
\end{CodeInput}
\end{CodeChunk}

@@

The ERA5 data \citep{hersbach2020era5} provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF) and can be downloaded from Copernicus Climate Data Store (CDS) \citep{cds} website or via the \pkg{ecmwfr} package \citep{ecwmfr}. This example reproduces the ERA5 data row of Figure 19 in \citet{hersbach2020era5} to show the southern polar vortex splitting into two on 2002-09-26 and further splitting into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also in \citet{simmons2020global} and \citet{simmons2005ecmwf}.

For the reproduction, we focus on the \code{era5-pressure} data, hourly pressure levels from 1970 to present, with the \emph{specific humidity} and \emph{geopotential}. The downloaded NetCDF data (\texttt{raw}) is first converted to a \code{cubble} object:

\begin{CodeChunk}
\begin{CodeInput}
R> dt <- as_cubble(
+   raw, vars = c("q", "z"),
+   long_range = seq(-180, 180, 1), lat_range = seq(-88, 88, 1))
\end{CodeInput}
\end{CodeChunk}

Creating the plot requires making transformations on time, unfolding the data for computing the statistic of interest, which is plotted directly as a contour plot with ggplot.

>>>>

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form
# spatial:  [144.8321, -37.98, 145.0964, -37.6655], Missing CRS!
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name              wmo_id ts               
  <chr>       <dbl> <dbl> <dbl> <chr>              <dbl> <list>           
1 ASN00086038  145. -37.7  78.4 essendon airport   95866 <tibble [10 x 4]>
2 ASN00086077  145. -38.0  12.1 moorabbin airport  94870 <tibble [10 x 4]>
3 ASN00086282  145. -37.7 113.  melbourne airport  94866 <tibble [10 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

The ERA5 reanalysis \citep{hersbach2020era5} provides hourly estimates of atmospheric, land and oceanic climate variables on a global scale. The hourly pressure level data in the NetCDF format can be downloaded from Copernicus Climate Data Store (CDS) \citep{cds} or via the \pkg{ecmwfr} package \citep{ecwmfr}. This example reproduces the row created from ERA5 reanalysis data shown in Figure 19 by \citet{hersbach2020era5}. The plot shows the southern polar vortex splitting into two on 2002-09-26, and further splitting into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also in \citet{simmons2020global} and \citet{simmons2005ecmwf}.

To work with NetCDF data in R, several packages are available, including \code{ncdf4}, \code{RNetCDF}, and \code{tidync}. The following code converts a NetCDF object of class \code{ncdf4} into a cubble object:

\begin{CodeChunk}
\begin{CodeInput}
R> raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))
\end{CodeInput}
\end{CodeChunk}

Analysts can extract a subset of the NetCDF data with the arguments \code{vars}, \code{long_range} and \code{lat_range}. In this example, the variables q (specific humidity) and z (geopotential) are selected and the coordinates are subsetted to every degree in longitude and latitude:

\begin{CodeChunk}
\begin{CodeInput}
R> (dt <- as_cubble(raw, vars = c("q", "z"),
+                  long_range = seq(-180, 180, 1), lat_range = seq(-88, -15, 1)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [26640], index: time, nested form
# spatial:  [-180, -88, 179, -15], Missing CRS!
# temporal: time [date], q [dbl], z [dbl]
     id  long   lat ts              
  <int> <dbl> <dbl> <list>          
1     1  -180   -15 <tibble [8 x 3]>
2     2  -179   -15 <tibble [8 x 3]>
3     3  -178   -15 <tibble [8 x 3]>
4     4  -177   -15 <tibble [8 x 3]>
5     5  -176   -15 <tibble [8 x 3]>
# i 26,635 more rows
\end{CodeOutput}
\end{CodeChunk}

Once the NetCDF data is coerced into a cubble object, subsequent analysis can be conducted to filter on the date of interest, scale the variable specific humidity and create visualisation in ggplot as in Figure \ref{fig:netcdf}.
